---
title: "dsci_II_project"
author: "Heather Lu"
date: "2026-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)      
library(caret)          
library(glmnet)         
library(mgcv)           
library(earth)          
library(randomForest)   
library(gbm)            
library(gtsummary)      
library(corrplot)       
library(ggplot2)        
```

```{r, include=FALSE}
load("datA.RData")
str(datA)
sum(is.na(datA))

# fix variable types
datA <- datA |> 
  mutate(
    menopause = as.factor(menopause), # 0 = Pre-menopausal, 1 = Post-menopausal
    race = as.factor(race), # 1 = White, 2 = Black, 3 = Asian, 4 = Hispanic 
    tumor_grade = as.factor(tumor_grade), # 1 = Well differentiated, 2 = Moderately differentiated, 3 = Poorly differentiated
    ER = as.factor(ER), # 0 = Negative, 1 = Positive 
    PR = as.factor(PR), # 0 = Negative, 1 = Positive
    HER2 = as.factor(HER2), # 0 = Negative, 1 = Positive
    stage = as.factor(stage), # 1 = Stage I, 2 = Stage II, 3 = Stage III
    surgery = as.factor(surgery), # 0 = Breast-conserving surgery, 1 = Mastectomy
    chemo = as.factor(chemo), # 0 = No, 1 = Yes
    hormonal_therapy = as.factor(hormonal_therapy) # 0 = No, 1 = Yes
  )
```
# Main Text

## Introduction / Background
...

## Methods


## Results
...

# Appendix

```{r}
# check outcome distribution 
hist(datA$log_PFS, main = "Distribution of log_PFS", xlab = "log_PFS")
```
*Figure 1. Distribution of log_PFS.*

```{r}
datA |> 
  select(age, bmi, tumor_size, lymph_nodes, Ki67) |> 
  cor() |> 
  corrplot(method = "number")
```
*Figure 2. Correlation matrix of continuous predictors in Dataset A.*

From the correlation plot, it seems that lymph_nodes and tumor_size has a moderate positive correlation r = 0.71 , which makes sense in a clinical context. Everything else seems largely uncorrelated; there is no severe multicollinearity, OLS shouldn't be too unstable. The mild correlation between lymph_nodes and tumor_size may result in one predictor being dropped by LASSO, so elastic net might be preferred here. 
