---
title: "dsci_II_project"
author: "Heather Lu"
date: "2026-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)      
library(caret)
library(tidymodels)
library(glmnet)         
library(mgcv)           
library(earth)          
library(gtsummary)      
library(corrplot)       
library(ggplot2)    
library(knitr)
```

# Load Data
```{r}
load("datA.RData")
str(datA)
sum(is.na(datA))

# fix variable types
datA <- datA |> 
  select(-id) |> 
  mutate(
    menopause = as.factor(menopause),
    race = as.factor(race),
    tumor_grade = as.factor(tumor_grade),
    ER = as.factor(ER),
    PR = as.factor(PR),
    HER2 = as.factor(HER2),
    stage = as.factor(stage),
    surgery = as.factor(surgery),
    chemo = as.factor(chemo),
    hormonal_therapy = as.factor(hormonal_therapy)
  )

```

# Table Summary datA
```{r}
table_a <- datA |> 
  select(-log_PFS) |>
  tbl_summary(
    label = list(
      age              ~ "Age at diagnosis (years)",
      bmi              ~ "BMI (kg/m^2)",
      tumor_size       ~ "Tumor size (cm)",
      lymph_nodes      ~ "Positive lymph nodes (n)",
      Ki67             ~ "Ki-67 index (%)",
      menopause        ~ "Menopausal status",
      race             ~ "Race/ethnicity",
      tumor_grade      ~ "Tumor grade",
      ER               ~ "Estrogen receptor status",
      PR               ~ "Progesterone receptor status",
      HER2             ~ "HER2 status",
      stage            ~ "Clinical stage",
      surgery          ~ "Surgery type",
      chemo            ~ "Chemotherapy",
      hormonal_therapy ~ "Hormonal therapy"
    ),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous()  ~ 1,
      all_categorical() ~ c(0, 1)
    )
  ) |> 
  modify_header(label = "**Characteristic**") |>
  modify_caption("**Table 1. Baseline characteristics of datA (1990–1995 cohort, n=2000)**") |>
  bold_labels()

table_a
```

# EDA
```{r}
# check outcome distribution 
hist(datA$log_PFS, main = "Figure 1. Distribution of log_PFS", xlab = "log_PFS")

```
*Figure 1\. Distribution of log_PFS.*

```{r}
datA |> 
  select(age, bmi, tumor_size, lymph_nodes, Ki67) |> 
  cor() |> 
  corrplot(method = "number")
```
*Figure 2\. Correlation matrix of continuous predictors in Dataset A.*

From the correlation plot, it seems that lymph_nodes and tumor_size has a moderate positive correlation r = 0.71 , which makes sense in a clinical context. Everything else seems largely uncorrelated; there is no severe multicollinearity, OLS shouldn't be too unstable. The mild correlation between lymph_nodes and tumor_size may result in one predictor being dropped by LASSO, so elastic net might be preferred here. 

# Train test split + model building 
```{r}
# train test split
set.seed(67)
data_split <- initial_split(datA, prop = 0.8)
training_data <- training(data_split)
testing_data <- testing(data_split)

# train control object
ctrl1 <- trainControl(method = "cv", number = 10, savePredictions = TRUE)
```

# OLS
```{r}
set.seed(67)

ols_fit <- train(log_PFS ~.,
                 data = training_data,
                 method = "lm",
                 trControl = ctrl1)

summary(ols_fit)
```
OLS results: 

- tumor_size: negative effect, larger tumors -> shorter PFS

- tumor_grade3: negative, poorly differentiated tumors -> shorter PFS

- lymph_nodes: negative, more positive nodes -> shorter PFS

- ER1: positive, ER positive -> longer PFS

- HER21: negative, HER2 positive -> shorter PFS

- hormonal_therapy1: positive, hormonal therapy -> longer PFS

- age, bmi, menopause, race, PR, stage, surgery, chemo not significant in OLS

- Adjusted R² = 0.20 : the model explains about 20% of variance in log PFS

# Ridge 
```{r}
set.seed(67)
ridge_fit <- train(log_PFS ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 0,
                                          lambda = exp(seq(6, -4, length=100))),
                   trControl = ctrl1)

plot(ridge_fit, xTrans = log)

ridge_fit$bestTune
```
optimal lambda is 0.125, small penalty. makes sense given we don't have severe multicollinearity, we don't need to shrink too much.

# LASSO
```{r}
set.seed(67)
lasso_fit <- train(log_PFS ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = exp(seq(4, -10, length=100))),
                   trControl = ctrl1)

plot(lasso_fit, xTrans = log)

lasso_fit$bestTune

coef(lasso_fit$finalModel, s = lasso_fit$bestTune$lambda)
```
LASSO dropped menopause, bmi, tumor_grade2, PR, stage3, surgery, chemo but kept age, race4, tumor_size, tumor_grade3, lymph_nodes, ER1, HER21, Ki67, stage2, hormonal_therapy1, race3. both lymph_nodes and tumor_size were kept, but shrunken compared to OLS.

# Elastic net

```{r}
set.seed(67)
enet_fit <- train(log_PFS ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                          lambda = exp(seq(4, -10, length=100))),
                   trControl = ctrl1)

enet_fit$bestTune
```
enet_fit best tuning parameters yielded alpha = 0.35, lambda = 0.0172. Model is leaning more toward ridge than lasso, which makes sense given the correlation between tumor_size and lymph_nodes. It's using the ridge component to handle those correlated predictors.

# GAM
```{r, warning=FALSE}
set.seed(67)

gam_fit <- train(log_PFS ~ .,
                 data = training_data,
                 method = "gam",
                 trControl = ctrl1)

gam_fit$bestTune
gam_fit$finalModel
```
- Applied smooth terms s() to the continuous predictors: lymph_nodes, tumor_size, bmi, Ki67, age

- s(Ki67) has edf = 4.32 — most nonlinear, suggesting Ki67 has a notably nonlinear relationship with log PFS

- s(lymph_nodes) edf = 1.44 and s(tumor_size) edf = 1.38 — close to 1, meaning nearly linear

s(bmi) and s(age) edf = 1.00 — essentially linear, no nonlinearity detected

# MARS
```{r}
set.seed(67)

mars_grid <- expand.grid(degree = 1:3, 
                         nprune = 2:15)

mars_fit <- train(log_PFS ~ .,
                 data = training_data,
                 method = "earth",
                 tuneGrid = mars_grid,
                 trControl = ctrl1)
ggplot(mars_fit)

mars_fit$bestTune

coef(mars_fit$finalModel) 
```
- degree = 1, no interactions needed

- optimal at nprune = 11

hinge functions: 
- h(8 - lymph_nodes) — effect of lymph nodes below 8

- h(tumor_size - 1.3) and h(1.3 - tumor_size) together — piecewise effect around tumor size = 1.3cm, a knot/breakpoint

- h(39 - Ki67) — effect of Ki67 below 39%

- h(23.1 - bmi) and h(bmi - 35.7) — nonlinear BMI effects at two breakpoints

- ER1, HER21, hormonal_therapy1 — linear terms

- h(tumor_size - 0.4) — another tumor size hinge

MARS confirms nonlinear effects for Ki67 and tumor size, consistent with the GAM findings. 

# resamples 

```{r}
set.seed(67)

resamp <- resamples(list(enet = enet_fit, ols = ols_fit, lasso = lasso_fit, rdige = ridge_fit, gam = gam_fit, mars = mars_fit))

summary(resamp)

parallelplot(resamp, metric = "RMSE")

bwplot(resamp, metric = "RMSE")
```
```{r}
cv_table <- summary(resamp)$statistics$RMSE |>
  as.data.frame() |>
  rownames_to_column(var = "Model") |>
  select(Model, Mean, Min., Max.) |>
  arrange(Mean) |>
  mutate(Model = Model,
         Mean = round(Mean, 4),
         Min. = round(Min., 4),
         Max. = round(Max., 4))

kable(cv_table, 
      col.names = c("Model", "Mean RMSE", "Min RMSE", "Max RMSE"),
      caption = "Table 2. 10-fold cross-validation RMSE for models fitted on Dataset A")
```

GAM has lowest median CV RMSE, followed by enet and lasso. All models overlap, differences are not significant. enet has one outlier fold and performed identically to LASSO, indicating the ridge component didn't change much. GAM captures the non-linear effects of Ki67 and tumor size, could potentially the best model. However, GAM is less interpretable, and the goal of the project is to understand how demographic, tumor, and treatment characteristics influence progression-free survival in breast cancer patients. 

no model is significantly better than any other,the differences in RMSE are not statistically significant. We will decide to go with LASSO over enet, since it only has 1 tuning parameter, no outliers and identical RMSE to enet.

# prediction
```{r}
lasso_pred <- predict(lasso_fit, newdata = testing_data)

# test RMSE
sqrt(mean((lasso_pred - testing_data$log_PFS)^2))

#R^2
cor(lasso_pred, testing_data$log_PFS)^2
```

test RMSE = 0.68, slightly lower than CV RMSE = 0.706. 

- 6 models trained and compared via 10-fold CV

- No significant differences between models

- Lasso selected as final model due to parsimony and interpretability

- Test RMSE = 0.684

- lasso model explains about 19% of variance in log PFS on the test set, consistent with the training CV R^2 of 0.20, which is a good sign that the model is not overfitting.

- final model: Lasso with lambda = 0.015, CV RMSE = 0.707, Test RMSE = 0.684,
test R^2 = 0.190

# Part II

```{r}
load("datB.RData")

datB <- datB |> 
  select(-id) |> 
  mutate(
    menopause = as.factor(menopause), # 0 = Pre-menopausal, 1 = Post-menopausal
    race = as.factor(race), # 1 = White, 2 = Black, 3 = Asian, 4 = Hispanic 
    tumor_grade = as.factor(tumor_grade), # 1 = Well differentiated, 2 = Moderately differentiated, 3 = Poorly differentiated
    ER = as.factor(ER), # 0 = Negative, 1 = Positive 
    PR = as.factor(PR), # 0 = Negative, 1 = Positive
    HER2 = as.factor(HER2), # 0 = Negative, 1 = Positive
    stage = as.factor(stage), # 1 = Stage I, 2 = Stage II, 3 = Stage III
    surgery = as.factor(surgery), # 0 = Breast-conserving surgery, 1 = Mastectomy
    chemo = as.factor(chemo), # 0 = No, 1 = Yes
    hormonal_therapy = as.factor(hormonal_therapy) # 0 = No, 1 = Yes
  )
```

# Comparison of datA and datB

```{r}
table_b <- datB |> 
  select(-log_PFS) |>
  tbl_summary(
    label = list(
      age              ~ "Age at diagnosis (years)",
      bmi              ~ "BMI (kg/m^2)",
      tumor_size       ~ "Tumor size (cm)",
      lymph_nodes      ~ "Positive lymph nodes (n)",
      Ki67             ~ "Ki-67 index (%)",
      menopause        ~ "Menopausal status",
      race             ~ "Race/ethnicity",
      tumor_grade      ~ "Tumor grade",
      ER               ~ "Estrogen receptor status",
      PR               ~ "Progesterone receptor status",
      HER2             ~ "HER2 status",
      stage            ~ "Clinical stage",
      surgery          ~ "Surgery type",
      chemo            ~ "Chemotherapy",
      hormonal_therapy ~ "Hormonal therapy"
    ),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous()  ~ 1,
      all_categorical() ~ c(0, 1)
    )
  ) |> 
  modify_header(label = "**Characteristic**") |>
  modify_caption("**Table 2. Baseline characteristics of datB (2000–2005 cohort, n=2000)**") |>
  bold_labels()

# merge side by side
tbl_merge(
  tbls = list(table_a, table_b),
  tab_spanner = c("**1990-1995 Cohort**", "**2000-2005 Cohort**")
) |>
  modify_caption("**Table 4. Baseline characteristics of Dataset A and Dataset B**")
```


# Predict with current LASSO
```{r}
# predict on datB
lasso_datB <- predict(lasso_fit, newdata = datB)

# RMSE on datB
sqrt(mean((lasso_datB - datB$log_PFS)^2))

#R^2
cor(lasso_datB, datB$log_PFS)^2

```

- RMSE increased substantially from 0.684 to 0.829, about a 21% increase in prediction error

- R^2 dropped from 0.190 to 0.157, explaining even less variance in the new cohort

```{r}
# Part II - datB analysis

# train/test split
set.seed(67)
data_splitB <- initial_split(datB, prop = 0.8)
training_dataB <- training(data_splitB)
testing_dataB <- testing(data_splitB)

# CV control
ctrl1 <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

# OLS
set.seed(67)
ols_fitB <- train(log_PFS ~ ., data = training_dataB, method = "lm", trControl = ctrl1)

# Ridge
set.seed(67)
ridge_fitB <- train(log_PFS ~ ., data = training_dataB, method = "glmnet",
                    tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(6, -4, length = 100))),
                    trControl = ctrl1)

# Lasso
set.seed(67)
lasso_fitB <- train(log_PFS ~ ., data = training_dataB, method = "glmnet",
                    tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(4, -10, length = 100))),
                    trControl = ctrl1)

plot(lasso_fitB, xTrans = log)

# Elastic net
set.seed(67)
enet_fitB <- train(log_PFS ~ ., data = training_dataB, method = "glmnet",
                   tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                          lambda = exp(seq(4, -10, length = 100))),
                   trControl = ctrl1)

# GAM
set.seed(67)
gam_fitB <- train(log_PFS ~ ., data = training_dataB, method = "gam", trControl = ctrl1)

# MARS
set.seed(67)
mars_fitB <- train(log_PFS ~ ., data = training_dataB, method = "earth",
                   tuneGrid = mars_grid, trControl = ctrl1)

# Model comparison
resampB <- resamples(list(ols = ols_fitB, ridge = ridge_fitB, lasso = lasso_fitB,
                          enet = enet_fitB, gam = gam_fitB, mars = mars_fitB))
summary(resampB)
bwplot(resampB, metric = "RMSE")
```
- CV RMSE range is 0.673-0.676, lower than datA 0.706-0.710

- GAM has the lowest mean RMSE (0.6731) again

- Ridge and enet are virtually identical (0.6737)

- All models very close again, we choose to compare LASSO models based on interpretability & consistency 

# New LASSO prediction part II
```{r}
set.seed(67)
# predict on datB test set
lasso_predB <- predict(lasso_fitB, newdata = testing_dataB)

# test RMSE
sqrt(mean((lasso_predB - testing_dataB$log_PFS)^2))

# R^2
cor(lasso_predB, testing_dataB$log_PFS)^2

coef(lasso_fitB$finalModel, s = lasso_fitB$bestTune$lambda)
```
# coefficient comparison

```{r}
# extract coefficients from both models
coef_A <- coef(lasso_fit$finalModel, s = lasso_fit$bestTune$lambda) |> 
  as.matrix() |> 
  as.data.frame()

coef_B <- coef(lasso_fitB$finalModel, s = lasso_fitB$bestTune$lambda) |> 
  as.matrix() |> 
  as.data.frame()

# combine into one table
coef_table <- cbind(coef_A, coef_B)
colnames(coef_table) <- c("datA (1990-95)", "datB (2000-05)")

coef_table

kable(coef_table, digits = 3, caption = "Table 1. Lasso coefficients from datA and datB models")
```
Changes 
- HER2 flipped from negative to positive. In the 1990s cohort HER2 positive was associated with worse PFS, but in the 2000s cohort it became slightly positive. HER2+ is no longer a poor prognostic factor. 

- Hormonal therapy effect strengthened: coefficient grew from 0.134 to 0.202, suggesting hormonal therapy became more effective or more appropriately prescribed.

- Lymph nodes effect weakened — from -0.062 to -0.030, potentially better treatments

- More variables retained in datB — suggesting the later cohort has more complex predictive relationships.
