---
title: "dsci_II_project"
author: "Heather Lu"
date: "2026-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)      
library(caret)
library(tidymodels)
library(glmnet)         
library(mgcv)           
library(earth)          
library(randomForest)   
library(gbm)            
library(gtsummary)      
library(corrplot)       
library(ggplot2)        
```

```{r, include=FALSE}
load("datA.RData")
str(datA)
sum(is.na(datA))

# fix variable types
datA <- datA |> 
  mutate(
    menopause = as.factor(menopause), # 0 = Pre-menopausal, 1 = Post-menopausal
    race = as.factor(race), # 1 = White, 2 = Black, 3 = Asian, 4 = Hispanic 
    tumor_grade = as.factor(tumor_grade), # 1 = Well differentiated, 2 = Moderately differentiated, 3 = Poorly differentiated
    ER = as.factor(ER), # 0 = Negative, 1 = Positive 
    PR = as.factor(PR), # 0 = Negative, 1 = Positive
    HER2 = as.factor(HER2), # 0 = Negative, 1 = Positive
    stage = as.factor(stage), # 1 = Stage I, 2 = Stage II, 3 = Stage III
    surgery = as.factor(surgery), # 0 = Breast-conserving surgery, 1 = Mastectomy
    chemo = as.factor(chemo), # 0 = No, 1 = Yes
    hormonal_therapy = as.factor(hormonal_therapy) # 0 = No, 1 = Yes
  )
```
# Main Text

## Introduction / Background
...

## Methods


## Results
...

# Appendix

# Load Data + EDA
```{r}
# check outcome distribution 
hist(datA$log_PFS, main = "Distribution of log_PFS", xlab = "log_PFS")
```
*Figure 1. Distribution of log_PFS.*

```{r}
datA |> 
  select(age, bmi, tumor_size, lymph_nodes, Ki67) |> 
  cor() |> 
  corrplot(method = "number")
```
*Figure 2. Correlation matrix of continuous predictors in Dataset A.*

From the correlation plot, it seems that lymph_nodes and tumor_size has a moderate positive correlation r = 0.71 , which makes sense in a clinical context. Everything else seems largely uncorrelated; there is no severe multicollinearity, OLS shouldn't be too unstable. The mild correlation between lymph_nodes and tumor_size may result in one predictor being dropped by LASSO, so elastic net might be preferred here. 

# Train test split + model building 
```{r}
# train test split
set.seed(67)
data_split <- initial_split(datA, prop = 0.8)
training_data <- training(data_split)
testing_data <- testing(data_split)

# train control object
ctrl1 <- trainControl(method = "cv", number = 10, savePredictions = TRUE)
```

# OLS
```{r}
set.seed(67)

ols_fit <- train(log_PFS ~. - id,
                 data = training_data,
                 method = "lm",
                 trControl = ctrl1)

summary(ols_fit)
```
OLS results: 

- tumor_size: negative effect, larger tumors -> shorter PFS

- tumor_grade3: negative, poorly differentiated tumors -> shorter PFS

- lymph_nodes: negative, more positive nodes -> shorter PFS

- ER1: positive, ER positive -> longer PFS

- HER21: negative, HER2 positive -> shorter PFS

- hormonal_therapy1: positive, hormonal therapy -> longer PFS

- age, bmi, menopause, race, PR, stage, surgery, chemo not significant in OLS

- Adjusted R² = 0.20 : the model explains about 20% of variance in log PFS

# Ridge 
```{r}
set.seed(67)
ridge_fit <- train(log_PFS ~ . - id,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 0,
                                          lambda = exp(seq(6, -4, length=100))),
                   trControl = ctrl1)

plot(ridge_fit, xTrans = log)

ridge_fit$bestTune
```
optimal lambda is 0.125, small penalty. makes sense given we don't have severe multicollinearity, we don't need to shrink too much.

# LASSO
```{r}
set.seed(67)
lasso_fit <- train(log_PFS ~ . - id,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = exp(seq(4, -10, length=100))),
                   trControl = ctrl1)

plot(lasso_fit, xTrans = log)

lasso_fit$bestTune

coef(lasso_fit$finalModel, s = lasso_fit$bestTune$lambda)
```
LASSO dropped menopause, bmi, tumor_grade2, PR, stage3, surgery, chemo but kept age, race4, tumor_size, tumor_grade3, lymph_nodes, ER1, HER21, Ki67, stage2, hormonal_therapy1, race3. both lymph_nodes and tumor_size were kept, but shrunken compared to OLS.

# Elastic net

```{r}
set.seed(67)
enet_fit <- train(log_PFS ~ . - id,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                          lambda = exp(seq(4, -10, length=100))),
                   trControl = ctrl1)

enet_fit$bestTune
```
enet_fit best tuning parameters yielded alpha = 0.35, lambda = 0.0172. Model is leaning more toward ridge than lasso, which makes sense given the correlation between tumor_size and lymph_nodes. It's using the ridge component to handle those correlated predictors.

# GAM
```{r}
set.seed(67)

gam_fit <- train(log_PFS ~ . - id,
                 data = training_data,
                 method = "gam",
                 trControl = ctrl1)

gam_fit$bestTune
gam_fit$finalModel
```
- Applied smooth terms s() to the continuous predictors: lymph_nodes, tumor_size, bmi, Ki67, age

- s(Ki67) has edf = 4.32 — most nonlinear, suggesting Ki67 has a notably nonlinear relationship with log PFS

- s(lymph_nodes) edf = 1.44 and s(tumor_size) edf = 1.38 — close to 1, meaning nearly linear

s(bmi) and s(age) edf = 1.00 — essentially linear, no nonlinearity detected
